---
title: "Data Viz Hackathon"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Goal

Students will be divided into teams. Each team will be assigned one of the following datasets and topics:

| Group | Dataset | Topic |
|------------|------------------------------|------------------------------|
|  | RCPA3 States dataset | crime |
|  | 538 Bechdel dataset | gender representation in movies |
|  | RCPA3 [**nes**](https://rdrr.io/cran/RCPA3/man/nes.html)(National Economic Survey) dataset | explaining Americans' happiness with life |
|  | RCPA3 [**nes**](https://rdrr.io/cran/RCPA3/man/nes.html)(National Economic Survey) dataset | News consumption |
|  | 538 [college_all_ages](https://www.rdocumentation.org/packages/fivethirtyeight/versions/0.6.2/topics/college_all_ages) dataset | employment and earnings by major |
|  | Titanic dataset | Who survived the Titanic? |

With the time allowed, each team is to create a series of visualizations.

After completion of the visualizations, all students will vote for the winning team (but can't vote for their own team) in the following categories:

1.  most truthful / functional
2.  most insightful / enlightening
3.  most beautiful / eye-catching

## Rules

1.  you may reuse code you find online (i.e. from R Graph Gallery) to help you build the components of your visualization (but must design and arrange the components yourself)
2.  you may apply code generated by a Large Language Model to build the components of your visualization - but all ideas and the organizations of these components should be your own.
3.  all students and the T.A. may vote (but cannot vote for their own team)

# Setup

```{r}
library(tidyverse)
```

# 1. Import Datasets

## 1.1 Import RCPA3 NES & States datasets

```{r}
# install.packages("RCPA3")
library(RCPA3)
data(package = "RCPA3")
```

```{r}
head(nes)
```

```{r}
head(states)
```

## 1.2 Import 538 datasets

```{r}
# install.packages("fivethirtyeight")
library(fivethirtyeight)
data(package = "fivethirtyeight")
```

```{r}
head(bechdel)
```

```{r}
head(college_all_ages)
```

## 1.3 Titanic dataset

```{r}
titanic <- as_tibble(Titanic)
```

[RCPA3 documentation](https://cran.r-project.org/web/packages/RCPA3/refman/RCPA3.html)

-   [**nes**](https://rdrr.io/cran/RCPA3/man/nes.html)(National Economic Survey) data

-   **states** dataset

-   **world** dataset

```{r}
# install.packages("RCPA3")
library(RCPA3)
data(package = "RCPA3")
```

[R Datasets](https://vincentarelbundock.github.io/Rdatasets/articles/data.html) (from multiple sources)

[Data is Plural](https://www.data-is-plural.com/) newsletter of "useful/curious" datasets

To view all installed datasets:

```{r}
data()
```

[Five Thirty Eight](https://cran.r-project.org/web/packages/fivethirtyeight/vignettes/fivethirtyeight.html) dataset

```{r}
# renv::install("fivethirtyeight")
# install.packages("fivethirtyeight")
library(fivethirtyeight)
data(package = "fivethirtyeight")
```

```{r}
help("fivethirtyeight")
```

```{r}
head(bechdel)
```

Fun datasets

1.  Bechdel dataset (538)

2.  States dataset: crimes

3.  NES dataset: news consumption

4.  NES dataset: views on diversity

5.  NES dataset: explaining Americans' happiness with life

6.  fpp3 dataset: Boston Marathon winning times

7.  HistData: Quarrels

8.  538: [**college_all_ages**](https://www.rdocumentation.org/packages/fivethirtyeight/versions/0.6.2/topics/college_all_ages)datasets

9.  538: russia_investigation (presidential investigations)

    ```{r}
    colnames(college_all_ages)
    ```
