---
title: "Data Viz Hackathon"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Goal

Students will be divided into teams. Each team will be assigned one of the following datasets and topics:

| Group | Dataset | Topic |
|------------|------------------------------|------------------------------|
|  |  |  |
|  | 538 Bechdel dataset | gender representation in movies |
|  | RCPA3 States dataset | crime |
|  | RCPA3 [**nes**](https://rdrr.io/cran/RCPA3/man/nes.html)(National Economic Survey) dataset | explaining Americans' happiness with life |
|  | RCPA3 [**nes**](https://rdrr.io/cran/RCPA3/man/nes.html)(National Economic Survey) dataset | News consumption |
|  | 538 [college_all_ages](https://www.rdocumentation.org/packages/fivethirtyeight/versions/0.6.2/topics/college_all_ages) dataset | employment and earnings by major |
|  | Titanic dataset | Who survived the Titanic? |

With the time allowed, each team is to create a series of visualizations with their scaffolding (title, axes labels, legends, labels, annotations, captions, etc.). Some recommended steps:

1.  identify the data variables you will use
2.  map the data variables to visual variables and choose plot types to represent your data
3.  add in highlights, callouts, scaffolding, text, etc. to help highlight patterns and tell stories with your data
4.  

After completion of the visualizations, all students will vote for the winning team (but can't vote for their own team) in the following categories:

1.  most truthful / functional
2.  most insightful / enlightening
3.  most beautiful / eye-catching

## Rules

1.  you may reuse code you find online (i.e. from R Graph Gallery) to help you build the components of your visualization (but must design and arrange the components yourself)
2.  you may apply code generated by a Large Language Model to build the components of your visualization - but all ideas and the organizations of these components should be your own.
3.  all students and the T.A. may vote (but cannot vote for their own team)

# Setup

```{r}
library(tidyverse)
```

# 1. Import Datasets

## 1.1 Import RCPA3 NES & States datasets

```{r}
# install.packages("RCPA3")
library(RCPA3)
data(package = "RCPA3")
```

```{r}
head(nes)
```

```{r}
head(states)
```

## 1.2 Import 538 datasets

```{r}
# install.packages("fivethirtyeight")
library(fivethirtyeight)
data(package = "fivethirtyeight")
```

```{r}
head(bechdel)
```

```{r}
head(college_all_ages)
```

## 1.3 Titanic dataset

```{r}
titanic <- as_tibble(Titanic)
```

```{r}
```
