---
title: "Creating a Shiny App with AI"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Creating a Shiny App with an AI Assistant

## **1. Project Overview & Learning Objectives**

### **Objective**

To design, build, and deploy a simple, interactive data visualization dashboard using **R Shiny**. R Shiny is an open-source web framework used to create visualization dashboards using R.

This project will synthesize your skills in data exploration, visualization design (ggplot2/plotly), storytelling, and basic web application development. You will create a dashboard that allows a user to explore a health-related dataset and gain insights into a specific question or topic.

### **A Key Component: The AI Copilot**

You are not only permitted but **encouraged** to use Large Language Models (LLMs) like ChatGPT, GitHub Copilot, or Claude to assist you in writing and debugging your Shiny code. A critical part of this assignment will be documenting and reflecting on how you used AI as a tool to accelerate your development and learning process.

### Guiding Framework

1.  AUTOMATE THE TEDIOUS: Let AI, more specifically LLMs, help you code and do much of the tedious work for you.
2.  FOCUS ON THE BIG IDEAS AND CONCEPTS:
3.  USE AI AS AN ASSISTANT NOT REPLACEMENT
4.  MULTI-SHOT PROMPTING - ENGAGE IN CONVERSATION
5.  USE AI AS A LEARNING ASSISTANT

## 2. Shiny Basics

To create a Shiny dashboard, there are n main steps:

1.  Create a sub-folder for your Shiny app
2.  Write the code for your app in file called `app.R` placed within your sub-folder.
3.  to run it locally, ...
4.  to host a Shiny app online, you have [a few different options](https://shiny.posit.co/py/get-started/deploy.html), but a common solution is to host it on the cloud using a [Posit Connect Cloud account](https://connect.posit.cloud/plans). With a free Posit Connect account you can create 5 unique applications, which may be active for up to a total of 20 hours per month. For more intensive uses, you'll have to purchase a plan or seek out another option.

[Shiny tutorial](https://shiny.posit.co/r/getstarted/shiny-basics/lesson1/) Lesson 1

**Install Shiny**

```{r}
# install.packages("shiny")

# OR:
# renv::install("shiny")
```

```{r}
library(shiny)
runExample("01_hello")
```

## 3. Building a Minimal Shiny App with AI

The main code for Shiny apps are stored in a file called `app.R`. Each dashboard should have its own folder / directory (therefore, only one `app.R` file should be found in this folder). To run a dashboard, run the following in the R Console: `runApp("dir_name")`.

Each `app.R` file has 3 main parts:

1.  a user interface object - controls the layout and appearance of your app
2.  a server function - instructions for building your app
3.  a call to the **shinyApp** function - creates Shiny app objects from a UI/server pair.

The outline of a very basic `app.R` file looks like this:

```         
library(shiny)
library(bslib)

# Define UI ----
ui <- page_sidebar(
 
)

# Define server logic ----
server <- function(input, output) {

}

# Run the app ----
shinyApp(ui = ui, server = server)
```

To build this into an app with some actual content, review Shiny [Lesson 2](https://shiny.posit.co/r/getstarted/shiny-basics/lesson2/). I recommend trying one or both of the following approaches:

1.  continue following the tutorials on the Shiny webpage, replacing their examples with your own data. Run each cell or chunk of code one at a time. Examine what it does.
2.  use a LLM chatbot or coding assistant to assist you in writing the necessary code to create a shiny app. To get this to work, describe your data in detail (including the exact name of the dataset and the column names) and give precise instructions about what you are trying to do.

You can take a look at the code I created with the assistance of Claude Sonnet 4.5 for a [basic shiny app](sample_shiny_app/app.R) (both `app.R` and `01_app-basic.R`) in the sample_shiny_app folder in this repo as well as two other, slightly more sophisticated apps.

## 4. Deploying your Shiny app to a webpage

1.  Create a free account with <https://www.shinyapps.io/>

    1.  note: this site will be deprecated in Jan 2027 and replaced by Posit Cloud Connect, so alternatively, you could sign up directly with Posit Cloud Connect, I just do not have any experience with this new platform yet. If you create a shinyapps account, however, your account and apps will be automatically migrated over to PCC when the transition occurs).

2.  Follow instructions on their webpage for deploying your account online. Each user of a free account is allowed 5 apps. You can use one of these for your project for this course and save the other four for future projects.
