---
title: "Setup"
author: "Jeremy Mikecz"
date: "2026-01-07"
format: html
editor: visual
---

------------------------------------------------------------------------

# Setup Instructions

This document provides detailed instructions for setting up your R environment for this course/workshop.

## Table of Contents

1.  [Installing R and RStudio](#installing-r-and-rstudio)
2.  [Downloading Course Materials](#downloading-course-materials)
3.  [Setting Up the R Environment](#setting-up-the-r-environment)
4.  [Verifying Your Setup](#verifying-your-setup)
5.  [Troubleshooting](#troubleshooting)

## Installing R and RStudio {#installing-r-and-rstudio}

### Install R

1.  Go to <https://cran.r-project.org/>
2.  Click on the link for your operating system (Windows, Mac, or Linux)
3.  Download and install the latest version (4.0 or higher)

**Windows users:** Download "base" R\
**Mac users:** Choose the appropriate .pkg file for your chip (Intel or Apple Silicon)\
**Linux users:** Follow distribution-specific instructions

### Install RStudio (Recommended)

1.  Go to <https://posit.co/download/rstudio-desktop/>
2.  Download RStudio Desktop (free version)
3.  Install using the downloaded installer

**Alternative:** You can use other R interfaces, but course instructions assume RStudio.

## Downloading Course Materials {#downloading-course-materials}

### Option 1: Using Git (Recommended for experienced users)

If you have Git installed:

```{bash}
git clone [REPOSITORY-URL]
cd [repository-name]
```

To update materials throughout the course:

```{bash}
git pull origin main

```

### Option 2: Download as ZIP

```         
 Click the green "Code" button on the repository page
 Select "Download ZIP"
 Extract the ZIP file to a memorable location (e.g., ~/Documents/R-Courses/)
 Important: To get updates, you'll need to re-download
```

## Setting Up the R Environment {#setting-up-the-r-environment}

This repository uses renv for package management, ensuring everyone has the same package versions.

### Step 1: Open the Project

Double-click the \[project-name\].Rproj file to open the project in RStudio.

### Step 2: Install renv

If prompted to install renv, click "Yes" or run:

```{r}
install.packages("renv")
```

### Step 3: Restore Packages

In the R console, run:

```{r}
renv::restore()
```

This will install all required packages. This may take 10-30 minutes depending on your system and internet connection.

Type `y` when prompted to proceed.

### Step 4: Test the Setup

Run the setup verification script:

```{r}
source("scripts/00_setup.R")
```

If you see "✓ Setup complete!" you're ready to go!

## Verifying Your Setup {#verifying-your-setup}

Run this code in your R console to verify everything is working:

```{r}
# Check R version
R.version.string

# Check if key packages load
required_packages <- c("tidyverse", "rmarkdown", "here")
sapply(required_packages, require, character.only = TRUE)

# Test plot generation
library(ggplot2)
ggplot(mtcars, aes(x = mpg, y = wt)) +
  geom_point() +
  labs(title = "Test Plot - If you see this, ggplot2 works!")
```

## Troubleshooting {#troubleshooting}

### Issue: "renv::restore() fails with compilation errors"

#### Solution 1: Install binary packages instead of compiling from source

```{r}
# On Mac/Linux, when prompted about installing from source, choose "No"
# Or set this option:
options(install.packages.check.source = "no")
renv::restore()
```

#### Solution 2: Install system dependencies (Linux users)

On Ubuntu/Debian:

```{bash}
sudo apt-get install libxml2-dev libcurl4-openssl-dev libssl-dev
```

### Issue: "Cannot open file 'data/raw/myfile.csv'"

#### Solution: Always use the here package for file paths

```{r}
library(here)
data <- read.csv(here("data", "raw", "myfile.csv"))
```

Using **setwd()** is not recommended \[EXPLAIN WHY!\]

### Issue: "Package 'X' is not available for R version Y"

#### Solution: Update R to the latest version, or use an older package version

```{r}
# Install specific older version
renv::install("package@version")
```

### Issue: RStudio doesn't recognize the project

#### Solution: Ensure you opened the `.Rproj` file, not just a script

-   Close RStudio

-   Navigate to the repository folder

-   Double-click the .Rproj file

### Issue: Git clone fails

#### Solution: Check your network connection, or try HTTPS instead of SSH

```{bash}
# If SSH fails, try HTTPS
git clone https://github.com/username/repository.git
```

### Still Having Issues?

\[insert possible steps, i.e. contact instructor, review particular online materials, etc.\]
